Density-dependent selection and the limits of relative fitness

### 2018-04-25 manuscript.tex
l

Adult mortality occurs after lottery recruitment at a constant, type-specific per-capita rate $d_i\geq 1$, and can affect adults recruited in the current iteration, such that the new abundance at the end of the iteration is $(n_i+\Delta_+ n_i)/d_i$ (Fig.~\ref{fig:lottery}). In terms of absolute fitness, this can be written as
\begin{equation}
W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). 
\end{equation}
Here $\frac{\Delta_+ n_i}{n_i}$ is the per-capita rate of territorial acquisition, and $1/d_i$ is the fraction of type $i$ adults surviving to the next iteration.

@@

Selection in our variable density lottery model is density-dependent. Since this a discrete time model, this means that when two types are present, the selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on $N$. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested. To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
\begin{equation}
\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), 
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. 
\label{eq:bdensitydependence}

It is clear $d$-selection is independent of density. On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype $1$ with equal frequencies. Here $b_1=1$, $d_1=0.5$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce. The wildtype equilibrium density is $n_1/T\approx 0.4$, suggesting that this population would rarely experience declining strength of $c$-selection with increasing $N$.}
\end{figure}

@@

We now turn to the issue of how density responds to selection, comparing our variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}. As we saw in the previous section, $c$-selection has no effect on population density in our variable-density lottery. To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Selection between types with identical equilibrium density but different inter-type competitive ability. (a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$. Trajectories do not follow the line $N=K_{11}=K_{22}$. (b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=0.1$ and $c_1/c_2=5$. Trajectories converge on the line $N=K_{11}=K_{22}$.}
\end{figure}

@@

By contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}. As a result, Eq.~\eqref{eq:canonical} holds in demographic equilibrium even though $c$-selection is density-dependent.

@@

For Eq.~\eqref{eq:canonical} to break down, the selection coefficient $s$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$. Eq.~\eqref{eq:canonical} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model. Thus, to threaten Eq.~\eqref{eq:canonical}, we require selection to be density-dependent, and also density to be changing. This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium. In this case the validity of Eq.~\eqref{eq:canonical} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonical} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection coefficients is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change. During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps. Note, however, that the selection coefficient for $b$-sweeps (as  defined by differences in $\Delta n_i/n_i$) does depend on frequency because of the $1/\overline{b}$ factor.

@@

In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

For simplicity, we introduce the notation $D=N/T$ and assume that $D$ is small. We can thus make the approximation $1-e^{-\overline{b}D}\approx \overline{b}D$ and $f(\overline{b},N)\approx \overline{b}(1-D)$. We expect this to be a conservative approximate based on the worst case scenario, because $N$ is most sensitive to an increase in $b$ in this low-density linear regime. We first calculate the value of $f(\overline{b},N)$ at the halfway point in a sweep, where the halfway point is estimated with simple linear averages for $b$ and $N$. The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $d_i=b_i(1-D_i)=b_j(1-D_j)$. We obtain
\begin{align}
f_{\rm{half}}=f(\frac{b_i+b_j}{2},\frac{N_i+N_j}{2})&=\frac{b_i+b_j}{2}\left(1-\frac{D_i+D_j}{2}\right) \nonumber \\
&=\frac{1}{4} (b_i+b_j)(2-D_i-D_j) \nonumber \\
&=\frac{1}{4} (2d_i+b_i(1-D_j)+b_j(1-N_i)).
\end{align}
Dividing by $d_i$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
\begin{align}
\frac{f_{\rm{half}}}{d_i}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
&=\frac{1}{4}\left(2+\frac{1}{1+\epsilon}+1+\epsilon\right)\nonumber \\
&=1+\frac{1}{4}(\epsilon^2-\epsilon^3+\ldots),
\end{align}
where we have used the Taylor expansion $\frac{1}{1+\epsilon}=1-\epsilon+\epsilon^2-\epsilon^3+\ldots$.

>>>>

Adult mortality occurs after lottery recruitment at a constant, type-specific per-capita rate $d_i\geq 1$, and can affect adults recruited in the current iteration, such that the new abundance at the end of the iteration is $(n_i+\Delta_+ n_i)/d_i$ (Fig.~\ref{fig:lottery}). In terms of absolute fitness, this can be written as
\begin{equation}
W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). \label{eq:absfit}
\end{equation}
Here $\frac{\Delta_+ n_i}{n_i}$ is the per-capita rate of territorial acquisition, and $1/d_i$ is the fraction of type $i$ adults surviving to the next iteration.

@@

Selection in our variable density lottery model is density-dependent. This means that when two types are present, the discrete-time selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} may depend on $N$. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because most territories are claimed without contest at low density, whereas at high density few unoccupied territories are available to be contested.

To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
\begin{equation}
\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is then clear that $d$-selection is independent of density, because the density-dependent factor $1 + \frac{\Delta_+ n_i}{n_i}$ in Eq.~\eqref{eq:absfit} is the same for types that differ in $d$ only. On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype $1$ with equal frequencies. Here $b_1=1$, $d_1=2$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce.}
\end{figure}

@@

In this section we discuss the response of density to selection, comparing our variable-density lottery to previous studies of density-regulated populations. Our focus will be on the competitive ability trait $c$, because, as mentioned in the previous section, $c$-selection has no effect on population density even though it affects fitness under crowded conditions. Since this behavior is quite different from selection on more commonly used crowding-sensitivity parameters like $K$ in the logistic model, we start by revisiting MacArthur's $K$-selection argument \citep{macarthur_1967}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Selection between types with identical equilibrium density but different inter-type competitive ability. (a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$. Trajectories do not follow the line $N=K_{11}=K_{22}$. (b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=1.1$ and $c_1/c_2=5$. Trajectories converge on the line $N=K_{11}=K_{22}$.}
\end{figure}

@@

By contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}. As a result, the density-independent Eq.~\eqref{eq:canonicaldisc} holds in demographic equilibrium even though $c$-selection is density-dependent.

@@

For Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_i-W_j)/\overline{W}$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$. Eq.~\eqref{eq:canonicaldisc} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model. Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require selection to be density-dependent, and also density to be changing. This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium. In this case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize how Eq.~\eqref{eq:canonicalcont} breaks down in a simpler continuous-time birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence} has the same value $d_i-1$ at the beginning and end of a $b$-sweep. During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.

@@

In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$-sweep. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

For simplicity, we introduce the notation $D=N/T$ and assume that $D$ is small. We can thus make the approximation $1-e^{-\overline{b}D}\approx \overline{b}D$ and $f(\overline{b},N)\approx \overline{b}(1-D)$. We expect this to be a conservative approximate based on the worst case scenario, because $N$ is most sensitive to an increase in $b$ in this low-density linear regime. We first calculate the value of $f(\overline{b},N)$ at the halfway point in a sweep, where the halfway point is estimated with simple linear averages for $b$ and $N$. The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $f_{\rm initial}=b_i(1-D_i)=d_i-1=f_{\rm final}=b_j(1-D_j)$. We obtain
\begin{align}
f_{\rm{half}}=f(\frac{b_i+b_j}{2},\frac{N_i+N_j}{2})&=\frac{b_i+b_j}{2}\left(1-\frac{D_i+D_j}{2}\right) \nonumber \\
&=\frac{1}{4} (b_i+b_j)(2-D_i-D_j) \nonumber \\
&=\frac{1}{4} (2(d_i-1)+b_i(1-D_j)+b_j(1-D_i)).
\end{align}
Dividing by $d_i-1$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
\begin{align}
\frac{f_{\rm{half}}}{d_i-1}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
&=\frac{1}{4}\left(2+\frac{1}{1+\epsilon}+1+\epsilon\right)\nonumber \\
&=1+\frac{1}{4}(\epsilon^2-\epsilon^3+\ldots),
\end{align}
where we have used the Taylor expansion $\frac{1}{1+\epsilon}=1-\epsilon+\epsilon^2-\epsilon^3+\ldots$.
